CODECHECK: An Open Science initiative for the independent execution of computations underlying research articles during peer review improves reproducibility

### 2021-02-08 codecheck-paper.tex
edited annotated cert

We have a curated list of 25 certificates available at
\url{https://codecheck.org.uk/register}.
Table~\ref{tab:register} lists the certificates completed to date
along with citations to the original papers.
These CODECHECKs fall into three themes:
(1) classic and current papers from computational neuroscience,
(2) COVID-19 modelling preprints, and
(3) GIScience.

The first theme was the initial set of papers used to explore the concept
of CODECHECK. 
The idea was to take very well known historical articles from a 
specific domain, in this case the work area of SJE, possibly 
as as a conversation starter for potential collaborating journals.
This intention was brought to the first CODECHECK (certificate number
\texttt{2020-001}) performed before publication on an article for the journal
\emph{GigaScience}, which successfully reproduced the visualisations,
though not the full machine learning process.

The second theme is rather opportunistic. As the COVID-19 pandemic swept
over the globe, we hoped to contribute a small part to science's answer
to the unknown virus. The checks were solicited through community interaction
%Several COVID-19 models have been done during the time of peer review of these papers,
or by our initiative rather than requested from journals, but some certificates
have since then been acknowledged in reviewed papers
\cite{Davies2020-vj,kucharski_effectiveness_2020}. Unlike many claims in
popular media at the time,
% reference missing here TODO SJE, you had a note about some Noorden news piece
we found the results of Report 9, a model of the UK lockdown procedures,
to be reproducible (certificate number \texttt{2020-10}).

The third theme represents DN's service as a Reproducibility Reviewer at the AGILE conference series, where the \emph{Reproducible AGILE} Initiative \cite{reproducible_agile} established a process for reproducing workflows at the AGILE conference series independent of CODECHECK \cite{nust_improving_2020}.
While using slightly different terms and infrastructure (``reproducibility
reports'' are published on the Open Science Framework (OSF) instead of certificates on Zenodo) the 
AGILE reproducibility reviews adhere to the CODECHECK principles.
Furthermore, a small number of checks were completed as part of peer reviews
for GIScience journals.

@@

At the completion of each CODECHECK, the codechecker writes a certificate stating which outputs from the original article, i.e.,~numbers, figures or tables, could be reproduced.
This certificate is made openly
available so that all readers of the article can see which elements
were reproduced and what limitations or issues were faced.
The certificate also includes links to the code
and data used by the codechecker, allowing others to build on the
work.
The format of the certificates has evolved slightly during the
project, as we have learnt to automate different aspects of the
certification, such as inserting figures into certificate documents
and managing core metadata surrounding a check.
The metadata is stored in a machine-readable structured file in YAML,
the CODECHECK configuration file \texttt{codecheck.yml}\footnote{
See technical specification of the CODECHECK configuration file at
\url{https://codecheck.org.uk/spec/config/latest/}.}.
The configuration file enables current and future automation of workflows and
meta-analyses.

To demonstrate the typical contents of a certificate and the underlying
metadata, 
Figure~\ref{fig:cert} shows the first four out of ten
pages of the certificate \texttt{2020-012} \cite{cert-2020-012} with
annotations.
The referenced article \cite{unwin_report_2020} describes the modelling of COVID-19 across the USA.
Figure~\ref{fig:cert}A shows the certificate number and its DOI,
which points to the certificate and any supplemental files on Zenodo.
The CODECHECK logo is
added for recognition and to denote successful reproduction.
Figure~\ref{fig:cert}B provides the key metadata as a table generated
from the \texttt{codecheck.yml}; in particular it names the paper that
was checked (title, DOI), the authors, the codechecker, when the check was performed,
and in which repository related material is
accessible.
Figure~\ref{fig:cert}C shows a textual summary of how the CODECHECK was
performed and any interesting findings.
Figure~\ref{fig:cert}D (page 2 of the certificate) shows a table that
lists the outputs that were generated from the check based on the 
manifest of output files in the CODECHECK.
The certificate's file table shows the file name, the description stating to
which figure/table each file should be compared in the original paper, and
the size of the file.
Page 3 of the certificate,
as Figure~\ref{fig:cert}E shows, provides detailed notes from the
codechecker. In this particular example, the notes document what steps were 
needed to initialize the environment and run the code and that the execution
took quite long to run -- about 17~hours. Finally, on page 4 of
the certificate, we see the first output that was generated by the
CODECHECK (Figure~\ref{fig:cert}F). In this case, the figure matches
figure~4 of \cite{unwin_report_2020}.
Subsequent pages of the certificate show other outputs and the computing
environment in which the certificate itself was created (not shown here).

>>>>

We have a curated list of 25 certificates available at
\url{https://codecheck.org.uk/register}.
Table~\ref{tab:register} lists the certificates
along with citations to the original papers.
These CODECHECKs fall into three themes:
(1) classic and current papers from computational neuroscience,
(2) COVID-19 modelling preprints, and
(3) GIScience.

The first theme was the initial set of papers used to explore the
concept of CODECHECK.  The idea was to take very well known historical
articles from a domains of interest to one of the co-authors (SJE).
Our first CODECHECK (certificate number \texttt{2020-001}) was
performed before publication on an article for the journal
\emph{GigaScience}, which visusalized the outputs from a family of
supervised classification algorithms.

The second theme was a response to the COVID-19 pandemic, choosing
modelling papers to predict outcomes. The checks were solicited
through community interaction or by our initiative rather than
requested from journals.  Some certificates were since acknowledged in
the accepted papers
\cite{Davies2020-vj,kucharski_effectiveness_2020}. In particular, we
codechecked the well-known Imperial college model of UK lockdown
procedures from March 2020(certificate \texttt{2020-10}).  Contrary to
opinion in the mainstream media, the model results were reproducible
(CITE NOORDEN).

The third theme represents co-author DN's service as a Reproducibility
Reviewer at the AGILE conference series, where the \emph{Reproducible
  AGILE} Initiative \cite{reproducible_agile} independently
established a process for reproducing workflows at the AGILE
conference series \cite{nust_improving_2020}.  While using slightly
different terms and infrastructure (``reproducibility reports'' are
published on the Open Science Framework (OSF) instead of certificates
on Zenodo) AGILE reproducibility reviews adhere to CODECHECK
principles.  A few checks were also completed as part of peer reviews
for GIScience journals.

@@

At the end of each CODECHECK, the codechecker writes a certificate
stating which outputs from the original article, i.e.,~numbers,
figures or tables, could be reproduced.  This certificate is made
openly available so that all readers of the article can see which
elements were reproduced and what limitations or issues were found.
The certificate links to code and data used by the codechecker,
allowing others to build on the work.  The format of the certificates
evolved during the project, as we have learnt to automate different
aspects of the certification, such as inserting figures into
certificate documents and managing metadata of a check.  The metadata
is stored in a machine-readable structured file in YAML, the CODECHECK
configuration file \texttt{codecheck.yml}\footnote{ See technical
  specification of the CODECHECK configuration file at
  \url{https://codecheck.org.uk/spec/config/latest/}.}.  The
configuration file enables current and future automation of workflows
and meta-analyses.

Figure~\ref{fig:cert} shows pages 1--4 (of 10) of a typical
certificate \texttt{2020-012} \cite{cert-2020-012}.  The referenced
article \cite{unwin_report_2020} described models of COVID-19 across
the USA.  Figure~\ref{fig:cert}A shows the certificate number and its
DOI, which points to the certificate and any supplemental files on
Zenodo.  The CODECHECK logo is added for recognition and to denote
successful reproduction.  Figure~\ref{fig:cert}B provides the key
metadata as a table generated from \texttt{codecheck.yml}; it names
the paper that was checked (title, DOI), the authors, the codechecker,
when the check was performed, and in which repository related material
is accessible.  Figure~\ref{fig:cert}C shows a textual summary of how
the CODECHECK was performed and any interesting findings.
Figure~\ref{fig:cert}D (page 2 of the certificate) shows the outputs
that were generated based on the MANIFEST of output files in the
CODECHECK.  It shows the file name (Output), the description stating
to which figure/table each file should be compared in the original
paper (Comment), and the file size.  Page 3 of the certificate,
(Figure~\ref{fig:cert}E) gives detailed notes from the codechecker,
here documenting what steps were needed to run the code and that the
code took about 17~hours to complete. Finally, page 4 of the
certificate shows the first output generated by the CODECHECK
(Figure~\ref{fig:cert}F). In this case, the figure matched figure~4 of
\cite{unwin_report_2020}.  The remaining pages of the certificate show
other outputs and the computing environment in which the certificate
itself was created (not shown here).
